A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-14 paper/paper.tex
fix unicode issue/bugs
fix terminology skew with how contexts are named in Fig 1

\paragraph{Basics of Profunctors.}
While we assume the reader has some background knowledge of category
theory, we briefly define profunctors, which are not included in many
introductory texts.  Recall that a category $\cat C$ has a collection of
objects and morphisms with identity and composition, and a functor $F :
\cat C \to \cat D$ is a function on objects and a function on morphisms
that preserves identity and composition.  A category can be thought of
as a generalization of a preordered set, which has a set of elements and
a binary \emph{relation} on its objects satisfying reflexivity and
transitivity.  A category is then a \emph{proof-relevant preorder},
where morphisms are the proofs of ordering, and the reflexivity and
transitivity proofs must identity and unit equations. A functor is then
a \emph{proof-relevant monotone function}.  Given categories $\mathcal
C$ and $\mathcal D$, a profunctor $R$ from $\mathcal C$ to $\mathcal D$,
written $R : \cat C \pto \cat D$ is a functor $R : {\cat C}^{op} \times
{\cat D} \to \Set$.  Because a profunctor outputs a $\Set$ rather than a
proposition, it is itself a \emph{proof-relevant relation}.  Thinking of
categories as proof-relevant preorders, functoriality says that the
profunctor is downward-closed in $\cat C$ and upward-closed in $\cat D$.
Given profunctors $R, S : \cat C \pto \cat D$, a homomorphism from
$R$ to $S$ is simply a natural transformation, which in the preordered
setting is simply an implication of relations.

@@

Finally we have several forms of context which are used in the theory.
The contexts $\Gamma$ of term variables with their types are as usual;
we write ${\Gamma \isadtctx}$ to indicate that a context is well-formed.
The category contexts $\Gamma \vdash \Xi \boundary$ contain object
variables with their categories; the two forms of category context are
$\alpha : \cat C$, containing one variable that can be used both
contravariantly and covariantly, and $\alpha : \cat C ; \beta : \cat D$,
containing a contravariant variable $\alpha$ and covariant variable
$\beta$.  Finally, we have contexts $\Phi$ of element variables with
their sets, alternating with those sets' object variables.  A typical
$\Phi$ has the shape
\[
\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \alpha_2 : \cat C_2,
x_2:R_2(\alpha_2,\alpha_3), \ldots, R_n(\alpha_n,\alpha_{n+1}), \alpha_{n+1} : C_{n+1}
\]
and represents the composition of the relations $R_1, R_2, R_3, \ldots,
R_n$.  We write $d^-(\Phi)$ for the first category variable in $\Phi$
(which we regard as the negative/contravariant position), $d^+(\Phi)$
for the last category variable in $\Phi$ (which we regard as the
positive/covariant position), and $\Phi_1 \jnctx \Phi_2$ for the append
of two category contexts where the last variable in $\Phi_1$ is equal to
the first variable in $\Phi_2$.  Formal inductive definitions are in the
appendix, but intuitively we have: 
\[
\begin{array}{ccl}
d^-(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_1 : \cat C_1\\
d^+(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_{n+1}:\cat C_{n+1}\\
(\Phi_1,\beta:\cat D) \jnctx (\beta:\cat D, \Phi_2) & = & \Phi_1, \beta : \cat D, \Phi_2
\end{array}
\]

@@

\begin{figure}[t]
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1
      \pipe \graphProf \alpha \beta P \pipe \negPresheaf {\cat C} \pipe \posPresheaf {\cat C}
      \\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& 1 \pipe P \times Q\\
      &&& \pipe \harrapp M a b \pipe \negPresheafAppPtoX a b \pipe \posPresheafAppPtoX a b\\
      \textrm{Elements} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M \mid \text{refl}_{M} \\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \mid \ldots \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Category Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Set Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P,\beta:\cat D\\
    \end{array}
  \end{mathpar}
  \caption{\vett{} Syntactic Forms}
  \label{fig:syntax}
\end{figure}

@@

\item Elements/natural transformations: ${\Gamma \pipe \Phi \vdash s :
  R}$, where ${\Gamma \vdash \Phi \isavectx}$ and $\Gamma\pipe
  \underline \Phi \vdash R \isaSet$.  A transformation $s$ has a context
  of transformation variable and a single set output. To be well formed,
  the context and set must be parameterized by the same contravariant
  and covariant object variables. To ensure this, we use a coercion
  operation $\underline \Phi$ that erases everything in the context but
  the left-most and right-most object variables, i.e. $\underline{\alpha
    : \cat C} = \alpha : \cat C$ and $\underline{\Phi} =
  d^-(\Phi);d^+(\Phi)$ otherwise.

@@

Next, in Figure~\ref{fig:set-connectives}, we cover the connectives for
the sets/profunctors, which classify elements/natural transformations
(the $\beta/eta$-rules are in the appendix).
%% %
%% The top of the figure includes the formation rules for tensor products
%% $\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
%% \alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
%% the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$, and
%% the cartesian unit and product types $1$ and $R \times S$.  
%% %
%% The remainder of the figure gives the introduction and elimination rules
%% for these sets .  
%% %
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$, which is well-formed when given two objects of the
same category.  
%
Its introduction and elimination rules are analogous to the usual rules
for equality in Martin-L\"of type theory.  The introduction rule is the
identity morphism (reflexivity) and the elimination rule is an induction
principle: we can use a term of $s : \punitinXfromYtoZ {\cat C} A B$ by
specifying the behavior when $s$ is of the form $\punitrefl \alpha$ in
the form of a continuation $\alpha. t$.  Like the $J$ elimination rule
for equality in Martin-L\"of type theory, $P$ must be ``fully general'',
i.e. well-typed for variables $\alpha$ and $\beta$; this is because for
distinct variables $\alpha$ and $\beta$, $\punitinXfromYtoZ{\cat
  C}{\alpha}{\beta}$ denotes the unit, which has a universal property,
but $\punitinXfromYtoZ{\cat C} A B$ denotes a restriction of a unit,
which in general does not.  Those familiar with linear
logic~\cite{polakawpfenning} might expect a more general rule, with the
continuation $t$ typed in a context $\Phi_l \jnctx \Phi_r$ and $\Phi_l
\jnctx \Phi \jnctx \Phi_r$ as the context of the conclusion of the rule.
Because of dependency, this is not necessarily well-formed in cases
where the endpoints $A$ and $B$ of $\punitinXfromYtoZ{\cat C} A B$ are
not fully general (i.e. not distinct variables).  However, the instances
of this more general rule that do type check are derivable from our more
restricted rule using right/left-hom types.

>>>>

\textbf{Basics of Profunctors.}  While we assume the reader has some
background knowledge of category theory, we briefly define profunctors,
which are not included in many introductory texts.  Recall that a
category $\cat C$ has a collection of objects and morphisms with
identity and composition, and a functor $F : \cat C \to \cat D$ is a
function on objects and a function on morphisms that preserves identity
and composition.  A category can be thought of as a generalization of a
preordered set, which has a set of elements and a binary \emph{relation}
on its objects satisfying reflexivity and transitivity.  A category is
then a \emph{proof-relevant preorder}, where morphisms are the proofs of
ordering, and the reflexivity and transitivity proofs must identity and
unit equations. A functor is then a \emph{proof-relevant monotone
  function}.  Given categories $\mathcal C$ and $\mathcal D$, a
profunctor $R$ from $\mathcal C$ to $\mathcal D$, written $R : \cat C
\pto \cat D$ is a functor $R : {\cat C}^{op} \times {\cat D} \to \Set$.
Because a profunctor outputs a $\Set$ rather than a proposition, it is
itself a \emph{proof-relevant relation}.  Thinking of categories as
proof-relevant preorders, functoriality says that the profunctor is
downward-closed in $\cat C$ and upward-closed in $\cat D$.  Given
profunctors $R, S : \cat C \pto \cat D$, a homomorphism from $R$ to $S$
is simply a natural transformation, which in the preordered setting is
simply an implication of relations.

@@

Finally we have several forms of context which are used in the theory.
The contexts $\Gamma$ of term variables with their types are as usual;
we write ${\Gamma \isadtctx}$ to indicate that a context is well-formed.
We name the remaining contexts after the judgements that they are used
by.  The set contexts $\Gamma \vdash \Xi \boundary$, which will be used
to type-check sets, contain object variables with their categories. The
two forms of set context are $\alpha : \cat C$, containing one variable
that can be used both contravariantly and covariantly, and $\alpha :
\cat C ; \beta : \cat D$, containing a contravariant variable $\alpha$
and covariant variable $\beta$.  Finally, the transformation contexts
$\Gamma \vdash \Phi \isavectx$ contain element variables with their sets, alternating with those
sets' object variables with their categories.  A typical $\Phi$ has the
shape
\[
\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \alpha_2 : \cat C_2,
x_2:R_2(\alpha_2,\alpha_3), \ldots, R_n(\alpha_n,\alpha_{n+1}),
\alpha_{n+1} : \cat C_{n+1}
\]
and represents the composition of the ``relations'' $R_1, R_2, R_3, \ldots,
R_n$.  We write $d^-(\Phi)$ for the first category variable in $\Phi$
(which we regard as the negative/contravariant position), $d^+(\Phi)$
for the last category variable in $\Phi$ (which we regard as the
positive/covariant position), and $\Phi_1 \jnctx \Phi_2$ for the append
of two category contexts where the last variable in $\Phi_1$ is equal to
the first variable in $\Phi_2$.  Formal inductive definitions are in the
appendix, but intuitively we have: 
\[
\begin{array}{ccl}
d^-(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_1 : \cat C_1\\
d^+(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_{n+1}:\cat C_{n+1}\\
(\Phi_1,\beta:\cat D) \jnctx (\beta:\cat D, \Phi_2) & = & \Phi_1, \beta : \cat D, \Phi_2
\end{array}
\]

@@

\begin{figure}[t]
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1
      \pipe \graphProf \alpha \beta P \pipe \negPresheaf {\cat C} \pipe \posPresheaf {\cat C}
      \\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& 1 \pipe P \times Q\\
      &&& \pipe \harrapp M a b \pipe \negPresheafAppPtoX a b \pipe \posPresheafAppPtoX a b\\
      \textrm{Elements} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M \mid \text{refl}_{M} \\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \mid \ldots \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Set Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Transformation Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P,\beta:\cat D\\
    \end{array}
  \end{mathpar}
  \caption{\vett{} Syntactic Forms}
  \label{fig:syntax}
\end{figure}

@@

\item Elements/natural transformations: ${\Gamma \pipe \Phi \vdash s :
  R}$, where ${\Gamma \vdash \Phi \isavectx}$ and $\Gamma\pipe
  \underline \Phi \vdash R \isaSet$.  A transformation $s$ has a context
  $\Phi$ of transformation variable and a single set output $R$. To be
  well formed, the context and set must be parameterized by the same
  contravariant and covariant object variables. To ensure this, we use a
  coercion operation $\underline \Phi$ from transformation contexts to
  set contexts that erases everything in the context but the left-most
  and right-most object variables ($\underline{\alpha : \cat C} = \alpha
  : \cat C$ and $\underline{\Phi} = d^-(\Phi);d^+(\Phi)$).

@@

Next, in Figure~\ref{fig:set-connectives}, we cover the connectives for
the sets/profunctors, which classify elements/natural transformations
(the $\beta/eta$-rules are in the appendix).
%% %
%% The top of the figure includes the formation rules for tensor products
%% $\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
%% \alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
%% the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$, and
%% the cartesian unit and product types $1$ and $R \times S$.  
%% %
%% The remainder of the figure gives the introduction and elimination rules
%% for these sets .  
%% %
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$, which is well-formed when given two objects of the
same category.  
%
Its introduction and elimination rules are analogous to the usual rules
for equality in Martin-L\"of type theory.  The introduction rule is the
identity morphism (reflexivity) and the elimination rule is an induction
principle: we can use a term of $s : \punitinXfromYtoZ {\cat C} A B$ by
specifying the behavior when $s$ is of the form $\punitrefl \alpha$ in
the form of a continuation $\alpha. t$.  Like the $J$ elimination rule
for equality in Martin-L\"of type theory, $P$ must be ``fully general'',
i.e. well-typed for variables $\alpha$ and $\beta$; this is because for
distinct variables $\alpha$ and $\beta$, $\punitinXfromYtoZ{\cat
  C}{\alpha}{\beta}$ denotes the unit, which has a universal property,
but $\punitinXfromYtoZ{\cat C} A B$ denotes a restriction of a unit,
which in general does not.  Those familiar with linear
logic~\cite{polakawpfenning} might expect a more general rule, with the
continuation $t$ typed in a context $\Phi_l \jnctx \Phi_r$ and $\Phi_l
\jnctx \Phi \jnctx \Phi_r$ as the context of the conclusion of the rule.
Because of dependency, this is not necessarily well-formed in cases
where the endpoints $A$ and $B$ of $\punitinXfromYtoZ{\cat C} A B$ are
not distinct variables.  However, the instances of this more general
rule that do type check are derivable from our more restricted rule
using right/left-hom types.
